A Practical Optional Type System for Clojure

### 2012-10-24 dissertation/design-choices.tex
Half of Rowan's corrections

Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
treatment of \lstinline|nil|.
It allows us to track whether particular bindings are possibly \lstinline|nil|
down conditional branches, and makes it practical to throw type errors 
where \lstinline|nil| is not allowed (see section \label{sec:OccurrenceTyping} for examples).
This is not practical in Java's type system because \emph{null} is not directly
expressible, which makes it impractical to error when a \emph{NullPointerException}
exception is possible: according to the type system, almost every operation
could involve \emph{null}!
Eliminating \emph{NullPointerException} exceptions is entirely the programmers
responsibility in Java.
Typed Clojure helps the programmer by statically proving certain misuses of \lstinline|nil| are impossible.

The design choice of potentially misusing \lstinline|nil| results in a type error
needs further work to be completely satisfying.
Certain cases of type inference should be improved to avoid redundant code
that ``pleases the type checker''.
For example, we currently do not know how to infer \lstinline|(filter identity coll)|
as returning a sequence of non-nil elements, for any sequence \lstinline|coll|.
(see section \ref{sec:filterneg}).
It is also possible for code interacting with Java to introduce unexpected \emph{null}
values. Currently, we let the programmer define a set of unenforced \emph{assumptions}
where \emph{null} is allowed in Java fields and methods (see section \label{sec:javainterop}).
Adding runtime assertions that enforce these assumptions would allow more
accurate errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
interfacing with untyped Racket code.

>>>>

Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
treatment of \lstinline|nil|.
It allows us to track whether particular bindings are possibly \lstinline|nil|
down conditional branches, and makes it practical to throw type errors 
where \lstinline|nil| is not allowed (see section \ref{sec:OccurrenceTyping} for examples).
This is not practical in Java's type system because \emph{null} is not directly
expressible, which makes it impractical to error when a \emph{NullPointerException}
exception is possible: according to the type system, almost every operation
could involve \emph{null}!
Eliminating \emph{NullPointerException} exceptions is entirely the programmers
responsibility in Java.
Typed Clojure helps the programmer by statically proving certain misuses of \lstinline|nil| are impossible.

The design choice of potentially misusing \lstinline|nil| results in a type error
needs further work to be completely satisfying.
Certain cases of type inference should be improved to avoid redundant code
that ``pleases the type checker''.
For example, we currently do not know how to infer \lstinline|(filter identity coll)|
as returning a sequence of non-nil elements, for any sequence \lstinline|coll|.
(see section \ref{sec:filterneg}).
It is also possible for code interacting with Java to introduce unexpected \emph{null}
values. Currently, we let the programmer define a set of unenforced \emph{assumptions}
where \emph{null} is allowed in Java fields and methods (see section \ref{sec:javainterop}).
Adding runtime assertions that enforce these assumptions would allow more
accurate errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
interfacing with untyped Racket code.

### 2012-10-24 dissertation/experiments.tex
Half of Rowan's corrections

The library had to be modified and simplified considerably to be able to type check.
Many existing conduits had troublesome variable-parameters, reminiscent of the
issues of assigning types to \lstinline|assoc| and \lstinline|partial|
(which is future work, see section \label{sec:futurearity}).
In these cases, simplified versions of the conduits were created that 
took fixed arguments.

>>>>

The library had to be modified and simplified considerably to be able to type check.
Many existing conduits had troublesome variable-parameters, reminiscent of the
issues of assigning types to \lstinline|assoc| and \lstinline|partial|
(which is future work, see section \ref{sec:futurearity}).
In these cases, simplified versions of the conduits were created that 
took fixed arguments.

### 2012-10-24 dissertation/intro.tex
Half of Rowan's corrections

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This is idea not new~\cite{Fag91,CF91,AM91}, but recent attempts~\cite{Tob10}
are noteworthy for their broad success in matching many of the advantages of statically typed
languages, notably due to the use of bidirectional checking~\cite{PT00}.
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

@@

\begin{itemize}
  \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
        occurrences of Java's \emph{null}. We make \emph{null} directly expressible
        in Typed Clojure's type system as \lstinline|nil|. This makes it reasonable for it to be a
        static type error if \emph{null} is potentially misused, an improvement over Java's static type system
        where \emph{null} cannot be expressed.

@@

  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
        which is able to type check many common Clojure idioms.
        We show how this type checker can be used to aid Clojure programmers use sophisticated
        programming styles by porting most of an existing library for monadic programming.

@@

  \item We identify major future issues in typing Clojure code. Multimethod definitions are problematic
        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
        Records have subtle semantics for dissociating keys, like dissociating a base key of a record returns
        a plain map.
%        Heterogeneous map types 
\end{itemize}

Together,

These contributions support my thesis by using the prototype to check practical code
and

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.
\footnote{F-BOUND EXAMPLE NEEDED}

>>>>

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This is idea not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
are noteworthy for their broad success in matching many of the advantages of statically typed
languages (see work by Tobin-Hochstadt~\cite{Tob10}), 
notably due to the use of bidirectional checking (like work by Pierce and Turner~\cite{PT00}).
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

@@

\begin{itemize}
  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
        which is able to type check many common Clojure idioms.
        We show how this type checker can be used to aid Clojure programmers use sophisticated
        programming styles by porting most of an existing library for monadic programming.

  \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
        occurrences of Java's \emph{null}. We make \emph{null} directly expressible
        in Typed Clojure's type system as \lstinline|nil|. This makes it reasonable for it to be a
        static type error if \emph{null} is potentially misused, an improvement over Java's static type system
        where \emph{null} cannot be expressed.

@@

@@

  \item We identify the main future issues in typing Clojure code. Multimethod definitions are problematic
        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
        Records have subtle semantics for dissociating keys, like dissociating a base key of a record returns
        a plain map.
\end{itemize}

Together these contributions support my thesis and motivate further work in this area.

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.
This would be required mainly for future work involving Java Generics, like typing
invocations of \emph{java.util.Collection/max}\footnote{\url{http://docs.oracle.com/javase/1.5.0/docs/api/java/util/Collections.html#max(java.util.Collection)}}.

### 2012-10-24 dissertation/lit-review.tex
Half of Rowan's corrections

Typed Clojure adds static type checking to an existing untyped language.
Coming from the other direction, Albardi, Cardelli, Pierce and Didier~\cite{ACPR94}
extend a static language to be more dynamic.
This work centres around the type \emph{Dynamic}, which represents
a static type in a dynamically typed languages.

Rossberg~\cite{Ros07} describes the advantages of having a type
\emph{Dynamic} over other approaches to making statically typed languages
more like dynamic ones. He says that type \emph{Dynamic} models
allows aspects of dynamic style to be represented in statically
typed languages without compromising the type system.
However, he claims type \emph{Dynamic} is rarely used in practical languages
because they are too inconvenient for many cases and they destroy
a valuable property of polymorphic, statically typed languages.

@@

Wright and Cartwright~\cite{WC97} developed Soft Scheme, a practical
soft type system for Scheme. 
It extended earlier work by Cartwright and Fagan~\cite{CF91}
and Fagan~\cite{Fag91}, adding support for practical features such as
first-class continuations and variable-arity functions.
Soft Scheme does not require any extra type annotations.

Type systems for Scheme have since moved away from soft typing.
Alas, the only reasonably complete reference for commentary on this transition 
appears to be slides from a talk by Felleisen~\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen says that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
error reporting that required PhD-level expertise to decipher.

@@

Intersection, union, and singleton types fit best as extrinsic types.
Hayashi~\cite{Hay91} describes two type theories, ATT which includes
intersection, union, and singleton types, and ATTT which further extends ATT
to include refinement types, which can be classified as extrinsic types.
Typed Clojure can also be considered an extrinsic type system which supports
union, intersection, and singleton types, but in less sophisticated forms
than ATT and ATTT.
In particular, singleton types in ATTT a much more advanced than Typed Clojure,
which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
and, unlike ATTT, no way to parameterise over singletons.

Several interesting projects have used intersection or union types.
Intersection types were originally introduced by Coppo, Dezani-Ciancaglini, and Venneri~\cite{CDV81}.
The use union types to type check dynamically typed languages with soft typing systems
dates back to Cartwright and Fagan~\cite{CF91}.

@@

Refinement types add a level of types refining an existing type system
in order to type check more detailed properties and invariants than standard static type systems
(described by Freeman and Pfenning~\cite{FP91}).
Refinement types are reminiscent of Typed Clojure, in that Typed Clojure
adds a level of types on top of Clojure.
Intersection types are critical here to allow more than one property or invariant
to be expressed for a function.
SML CIDRE is a refinement type checker for Standard ML by Davies~\cite{Dav05}.

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket~\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach for its representation of functions.
Typed Racket uses (disjoint) union types.

@@

Typed Racket fully expands all macro calls before type checking~\cite{Tob10} to
avoid the complex semantics of type checking macro definitions, an ongoing research area summarised
 by Herman~\cite{Her10}.
The design of my Clojure type system will follow a similar strategy; only the fully macro-expanded form
will be type checked. Type checking macro definitions are outside the scope of this project, and would
be exceptionally hard.

Along with a full static type system, Typed Racket 
also uses runtime contracts to enforce type invariants at runtime~\cite{TF08}.
Utilising runtime contracts to aid type checking is outside the scope of this project, but would be 
considered desirable and accessible future work.

@@

After some investigation, typing multimethods with Typed Clojure is assigned as future work
(see section \label{sec:mmfuture}).

@@

F-$\omega$ is a typed $\lambda$-calculus with support for higher kinds,
specifically type constructors (see Pierce~\cite{Pie02}).
Haskell~\cite{Mar10} distinguishes between types and type constructors,
the latter of which is an essential part of its monad library 
(see section \ref{sec:monads} for a monad library ported to Typed Clojure
that uses type constructors similarly).

>>>>

Typed Clojure adds static type checking to an existing untyped language.
Coming from the other direction, Abadi, Cardelli, Pierce and R\'{e}my~\cite{ACPR94}
extend a static language to be more dynamic.
This work centres around adding a type \emph{Dynamic}, which represents
a dynamic type in a statically typed language.

Rossberg~\cite{Ros07} describes the advantages of having a type
\emph{Dynamic} over other approaches to making statically typed languages
more like dynamic ones and surveys work in this area. 
He says that type \emph{Dynamic}
allows aspects of dynamic style to be represented in statically
typed languages without compromising the type system.
However, he claims type \emph{Dynamic} is rarely used in practical languages
because it is too inconvenient in real programs.
He claims to have fixed these issues, but a practical language has yet to integrated 
his changes.

@@

Wright and Cartwright~\cite{WC97} developed Soft Scheme, a
soft type system for Scheme. 
It extended earlier work by Cartwright and Fagan~\cite{CF91}
and Fagan~\cite{Fag91}, adding support for practical features such as
first-class continuations and variable-arity functions.
Soft Scheme does not require any extra type annotations.

Type systems for Scheme have since moved away from soft typing.
Alas, the only reasonably complete account of this transition 
appears to be slides from a talk by Felleisen~\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen comments that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
error reporting that required PhD-level expertise to decipher.

@@

Intersection, union, and singleton types are best considered as extrinsic types.
Hayashi~\cite{Hay91} describes two type theories, ATT which includes
intersection, union, and singleton types, and ATTT which further extends ATT
to include refinement types, which can be classified as extrinsic types.
Typed Clojure can also be considered an extrinsic type system which supports
union, intersection, and singleton types, but in less sophisticated forms
than ATT and ATTT.
In particular, singleton types in ATTT a much more advanced than Typed Clojure,
which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
and, unlike ATTT, no way to parameterise over singletons.

Several interesting projects have used intersection or union types.
Intersection types were originally introduced by Coppo, Dezani-Ciancaglini, and Venneri~\cite{CDV81}
for the $\lambda$-calculus.
The use of union types to type check dynamically typed languages with soft typing systems
dates back to Cartwright and Fagan~\cite{CF91}.

@@

Refinement types add a level of extrinsic types refining an existing intrinsic type system
in order to type check more detailed properties and invariants than standard static type systems
(described by Freeman and Pfenning~\cite{FP91}).
Refinement types have similarities with Typed Clojure, in that Typed Clojure
adds a level of extrinsic types on top of Clojure.
Intersection types are critical here to allow more than one property or invariant
to be expressed for a function.
SML CIDRE is a refinement type checker for Standard ML by Davies~\cite{Dav05}.

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket~\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach for its representation of functions.
Typed Racket uses union types which are used in Typed Clojure in very similar ways,
like for creating ML-like ``datatypes''.

@@

Typed Racket fully expands all macro calls before type checking~\cite{Tob10},
avoiding the complex semantics of type checking macro definitions, an ongoing research area summarised
 by Herman~\cite{Her10}.
Typed Clojure follows a similar strategy; only the fully macro-expanded form
will be type checked. Type checking macro definitions are outside the scope of this project.

Along with a full static type system, Typed Racket 
also uses runtime contracts to enforce type invariants at runtime
at the interface with untyped code ~\cite{TF08}.
Utilising runtime contracts to aid type checking is outside the scope of this project, but would be 
considered desirable and accessible future work.

@@

After some investigation, typing multimethods with Typed Clojure is assigned as future work
(see section \ref{sec:mmfuture}).

@@

F$_\omega$ is a typed $\lambda$-calculus with support for higher kinds,
specifically type constructors (see Pierce~\cite{Pie02}).
Haskell~\cite{Mar10} distinguishes between types and type constructors,
the latter of which is an essential part of its monad library 
(see section \ref{sec:monads} for a monad library ported to Typed Clojure
that uses type constructors similarly).
